\documentclass{article}
%\textwidth 11,8 cm
%\textheight 17 cm
\textheight 23 cm
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{stmaryrd}
\usepackage{makeidx}
\usepackage{times}
%Uncomment next line for pdflatex and use includegraphics with eps file
% for latex2html don't use the option [width=\textwidth]
% check that xfig files are exported magnif 100%
\usepackage{ifpdf}
\ifpdf
 \usepackage[pdftex]{hyperref}
\else
%\usepackage[ps2pdf,breaklinks=true,colorlinks=true,linkcolor=red,citecolor=green]{hyperref}
 \usepackage{pstricks}
\fi
\usepackage{graphicx}

\usepackage[latin1]{inputenc}
\usepackage[T1]{fontenc}
%\usepackage[francais]{babel}
\usepackage{latexsym}

\newcommand{\R}{{\mathbb{R}}}
\newcommand{\C}{{\mathbb{C}}}
\newcommand{\Z}{{\mathbb{Z}}}
\newcommand{\N}{{\mathbb{N}}}
\newcommand{\xcas}{{\tt xcas }}
\newcommand{\asinh}{\,\,\mbox{asinh\,}}
\newcommand{\atanh}{\,\,\mbox{atanh\,}}

\title{An introduction to the {\tt Xcas} interface}
\author{Bernard Parisse, University of Grenoble I}
\date{}
\makeindex
\begin{document}

\maketitle

\copyright 2007, 2009 Bernard Parisse. \\
Released under the Free Documentation License, 
as published by the Free Software Fundation.

\begin{abstract}
This document describes the Xcas software, which combines 
a computer algebra system, dynamic geometry, and a spreadsheet. 
It explains how to organize your worksheet. It does not
explain in detail the functions and syntax of the Xcas 
computer algebra system (see the relevant documentation).
\end{abstract}

\section{A first session}
To run \xcas, 
\begin{itemize}
\item under Windows\\
you must select the xcasen program in the Xcas program group,
\item under Linux\\
type \xcas in a terminal (xterm, rxvt, ...), 
or from a Gnome session, select Xcas in the Education menu,
\item under Mac OS X.4, 5\\
click on the \xcas icon in Applications.
\end{itemize}

You should see a new window with a menubar on the top (the main
menubar), a tab, a line with buttons (including a status button)
and a large space for this session. The cursor should
be in what we call the first level of the session; that is, in 
the white space at the right of the number 1 just
below the session menubar (otherwise click in this white space).
Type, for example, {\tt 30!} then hit the return key. 
You should see the answer and a new white space (with a number
2 at the left) ready for another entry.
Try a few more operations; e.g., type {\tt 1/3+1/6}
and hit return, or use the {\tt Graphic} menu, {\tt Curves} then
{\tt plotfunc}, etc.

You should now have a session with a few numbered pairs of
input/answers, each pair is named a level. The levels created so far
are all command lines (a command line is where you
type a command following the computer algebra system syntax).
Inside Xcas, there are other kinds of levels, the most frequently used
being comments, programs, geometry (2-d or 3-d) and spreadsheet levels.

The most important editing operations are:
\begin{itemize}
\item
Using the main menubar to enter an Xcas function name, for example\\
{\tt Cmds->Integer->is\_prime}\\
A dialog box will appear
with short help on the function, some examples and related
commands.
\item Modifying an existing input field and reevaluating it after
it has been modified by hitting the return key.
\item
Selecting a level or a group of levels with the mouse.
Selected level(s) may be moved or erased.
\item Using copy-paste, which follows Windows rules (Ctrl-C copy, Ctrl-X cut,
Ctrl-V paste) or Linux rules (select with the mouse to copy, paste with
the middle mouse button or with the paste button at the bottom).
\end{itemize}

Online help may be accessed using the main menubar {\tt Help->Index}
(short help, click on {\tt Details} for more help inside your browser) 
and {\tt Help->Manuals}. 
If you know a function name, you can also type the beginning and hit the
tab key (at the left of the Q key on a qwerty keyboard).

You can save your work using the {\tt Save} button or the
{\tt File} menu; this menu provides
other disk and printing operations.

The software configuration parameters are grouped in three sections:
Cas (computer algebra system), Graph (geometry and graphics), and General
(everything else). The main menubar {\tt Cfg} items will help
you modify the configuration (for the Cas configuration you may
also hit the status button, in the middle of the buttons line at 
the top of the session).

The buttons at the top of the session, just below the tab are:
\begin{itemize}
\item at the left, the help button,
\item in the middle left, the Save button for quick session saving,
\item in the middle, the status button 
(displaying Config followed by the number of digits,
the angle unit, etc.), you can click on it to modify
the cas configuration,
\item in the middle right, a STOP button to interrupt a computation, 
\item at the right, a Kbd button that makes
a scientific keyboard appear or disappear.
\item at the far right, a \framebox{x} that closes the session.
\end{itemize}

In the next sections, we will explain in more detail the
capabilities of the Xcas interface: online help,
the different kinds of levels,
the management of sessions and the configuration.

\section{Online help}
If you know a command name but want to check the syntax, type the 
beginning of the command name, then the tab key.  This will
display the command index, with the current position at the first
command after the command name beginning.  You get a short description
and a few examples displayed.  Left-clicking on an example should
complete the command line with this example. Right-clicking will
fill up the arguments inputlines with the example, so that you
can modify some of these arguments before clicking OK.
You can get a more detailed help inside your browser 
by clicking on the Details button (this works for most but not all commands).

If you don't know a command name, you might find it using the
menus or browsing the table of contents of the CAS manual. If you
know the TI89/TI Nspire CAS or the Maple language, you can 
also try their command names.

\section{Level types}
Each session is a numbered collection of levels, each level 
has a type: command line, comment, expression editor,
spreadsheet, program, 2-d or
3-d geometry, or logo-like turtle.

\subsection{Command line}
If no command line is available, you can get a new one by the 
\verb|Toolbox->New entry|
menu. You can modify existing command lines or write in a new
command line. The following shortcuts are available.
\begin{itemize}
\item Tab: show completions.
\item Enter: parse and evaluate the current command line. You can
type Enter while the cursor is at any position. A new command line
will be added if it was the last level of the session or if
the next level is not a command line.
\item Shift-Enter: make a new line in the current command line.
\item Ctrl-C: copy the selection to the clipboard.
\item Ctrl-V: paste the clipboard contents.
\item Ctrl-X: cut and copy the selection to the clipboard.
\item Ctrl-U: erase the command line without copying to the clipboard.
\item Esc: erase the command line and copy it to the clipboard.
\item Ctrl-Z: cancel the last edit command.
\item Up and down: move in the current command line if it is a
  multiline command line.
\item PageUp and PageDown: move to the previous or next level.
\item Ctrl-Up and Ctrl-Down: replace the selection with the previous
or next command that has been validated in a previous command line
(history saving mechanism).
\end{itemize}
When you evaluate a command line, a log output may be displayed
(written in blue over white by default), followed by the answer, which
might be an expression or a graphic (2-d or 3-d) widget if the
answer is a graphic object (for lists, the widget type corresponds
to the type of the last evaluated object).

If you do not want to see the answer displayed (for example if
you store a large matrix in a variable), just finish
your command line with \verb|:;|

\subsection{Comments}
Comments are inserted using the \verb|Toolbox->New comment| menu.
They are by default written in green instead of red over white.
Comments are not evaluated when pressing Enter (the effect is to
put the cursor to the next level).

\subsection{Expression editor} \label{sec:eqw}
Typing the shortcut {\tt Alt-e} 
or using the menu {\tt Expression->New expression} will
open an expression editor above the current level. The scientific
keyboard should automatically appear (press the \framebox{x} button at the 
top right of the keyboard if you don't want it).

The expression editor lets you enter complex mathematical expression
in 2-d like on a paper. 
For example, suppose you want to enter
\[ \left(\frac{x+1}{x+2}\right) \cdot \left(3+\frac{4}{x} \right)^5 \]
Open an expression editor ({\tt Alt-e}) then type
\begin{itemize}
\item $x+1 \ \uparrow$, this will select $x+1$
\item \verb|/| (for the division)
\item $x+2 \ \uparrow \ \uparrow$
\item
* 3 $\uparrow$ + 4 $\uparrow$ / x
\item
$\uparrow \ \uparrow \ \uparrow$ \verb|^| 5
\end{itemize}
You can now move the selection in the tree of the expression
with the arrows keys~:
\begin{itemize}
\item $\uparrow$ moves up one level in the tree, that is select the
function or operator with the current selection as argument 
(it also selects the other arguments of the function/operator),
\item $\downarrow$ moves down one level in the three, that is select
the first argument of the function/operator of the current selection,
\item $\leftarrow$ or $\rightarrow$ selects the left or right sibling 
of the current selection,
\item Ctrl-$\leftarrow$ or Crtl-$\rightarrow$ exchanges the current selection
with its left or right sibling.
\end{itemize}

Another example, let's enter
\[ \int_0^{+\infty} \frac{1}{1+x^2} \]
Type {\tt Alt-e} to open a new expression editor.
\begin{itemize}
\item Use the $\int$ key on the scientific keyboard or {\tt Ctrl-R} on
the keyboard to get an indefinite integral.
\item Use the \verb|,| key to switch to a definite integral.  (To do the
reverse operation, you can select each bound and press the backspace
key).
\item Use $\downarrow$ to replace 0 by the function \verb|1/(1+x^2)|.

\item Use $\uparrow$ and $\leftarrow$ to select the upper bound.
\item Use the $\infty$ key on the scientific keyboard to replace the 1.
\item Use the Enter key to evaluate the integral.
\end{itemize}

We now list the other special keys in the expression editor:
\begin{itemize}
\item \verb|"|: 
Outside editor mode (no visible cursor), 
edit the selection like in a command line, this is especially
useful if you want to edit a function name.
In editor mode, switch between considering the edit as a string
and as a normal edit. 
\item \verb|,|: if a sequence is selected, add a 0 to the end of the
sequence and select it. If the selection is not a sequence, creates
a sequence of size 2 with first element the selection, init the second
element to 0 and select it. You can move the newly created 0 in the
sequence with Ctrl-$\rightarrow$ and Ctrl-$\leftarrow$.
\item \verb|(|: if a sequence is selected, make it the arguments to a
user function, the name of the user function is initialized
to \verb|f| and selected (type the function name to replace 
the \verb|f| by your function name). If a vector is selected, it is replaced
by a sequence.
\item \verb|[|: if a sequence is selected, it is replaced by a vector,
if a vector is selected, an indexed vector/matrix position is created
where the variable name of the vector/matrix is initialized to
\verb|m| and selected (type the matrix/list name to replace \verb|m|).
\item Keys in combination with the Ctrl key:
\begin{itemize}
\item Ctrl-A: select All.
\item Ctrl-C: copy the selection to the clipboard (like under windows).
\item Ctrl-V: paste the selection from the clipboard.
\item Ctrl-T: translate the selection to \LaTeX\ and copy it to the clipboard
(the translation is also copied to the \verb|Msg| output, click
on the \verb|Msg| button to make it visible or hide it).
\item Ctrl-E: evaluate the selection.
\item Ctrl-F: factor the selection.
\item Ctrl-N: apply normal to the selection.
\item Ctrl-P: apply partfrac to the selection.
\item Ctrl-I (or Tab key): completion key (open the position in the index at the function
whose name begins the same).
\item Ctrl-R: integral.
\item Ctrl-S: simplify the selection.
\item Ctrl-L: limit.
\item Ctrl-D: derivative.
\item Ctrl-Z: undo.
\item Ctrl-Y: redo.
\end{itemize}
\end{itemize}

\subsection{Programs}
Programs may be written in a command line if they are one or two
lines long, but for more complex programs, it is a good idea to
put them in a separate program level, using the
\verb|Prg->New program| menu. This will open an editor in
a new level. Type your program (the {\tt Prg} menu may help you
for the syntax of tests, loops, function definition, etc.), 
and press the {\tt OK} button once completed. 
If parsing is successful, you can test your program
in any command line.

The \verb|Prg| menu has items to open or
import an existing (program) file inside the current text, save
or export the current program (independently from the worksheet). 
You can also use the menu 
to insert programming structures. There are buttons to find the
next occurence of a search string, to parse the current program
(errors are displayed in the message area) and to save the
program (the current filename is displayed at the right of the
save button).

The keywords of the language are highlighted. In Xcas programming
mode, the tab key may be used for indentation (experimental
feature).

Tip: before pressing the {\tt OK} button, check the programming
syntax by looking at the status area button
in the middle at the top of the session. If you need
to change the syntax, hit the status button and select the right syntax.

\subsection{Geometry and graphics}
A graphic output is made of a 2-d or 3-d scene with controls,
it corresponds to a unique command line above it. 
You can create some examples easily using the {\tt Graphic} menu.

A geometry figure is the display of a 2-d or 3-d graphic scene 
representing the graphical objects of the command lines 
that are on its left.
\begin{itemize}
\item You can create some geometric objects using the mouse (points,
lines, triangles, circles, etc.).
\item Alternatively you can enter command lines, with the help
of the Geo menu.
\item
Interactive geometry is implemented by moving points in pointer
mode or by moving special objects called parameters or cursors. 
\end{itemize}

\subsubsection{Common features}
The controls are common to 2-d, 3-d graphics or geometry. You can
move left, right, up and down the window, redefine it using
coordinates, zoom in and out, pause and restart an animation using the
buttons and the {\tt M} menu at the right of the graphic, or 
using the {\tt Graphic} menu.
In 2-d, the coordinates 
of the mouse are displayed while it is in the graphic area. 
You can modify the visualization
window with the mouse (drag to move, 
use the scroll button for centered zoom in/out, or select an area with
the right mouse button for an area zoom in).
In 3-d,
you can rotate the representation along the current x, y, z axis
with a mouse drag outside of the scene rendering, or
by using the x,X, y,Y and z,Z keys (if the focus is in the 3-d scene).

Note that the controls apply only to the visualization, not to the creation
of graphic objects. For example, if you create a plot of a function,
the discretisation parameters for this plot (e.g. minimal value for
$x$) are fixed and will not change if you move the minimal visualisation
value using the controls. You must recreate a graphical object if
you did not correctly choose the parameters. For example if you
run the command \verb|plot(sin(x))|, the minimal value for computation
of $x$ will be the default value 
(Xmin from the Geo configuration, -10 by default). If
you move the minimal value of $x$ for visualization below Xmin, you will
not see the plot there. To see the plot below Xmin, you would have
to specify something like \verb|plot(sin(x),x=-20..0)| and validate
the command line. This limitation does not apply to geometric objects
(points, segments, lines and circles) 
which will always be drawn if they are visible.

An {\tt M} menu is available below the controls (its items are also available
from the top {\tt Graphic} menu), 
it contains items to print
and control the screen. You can export a graph to encapsulated
postscript and to PNG (you must have the converting tools 
netpbm or ImageMagic
installed, this should be installed by the Xcas installer under Windows. If they
are not installed under Linux, run your software installer or 
open a Terminal
and type \\
\verb|sudo apt-get install imagemagick| \\
for Debian/Ubuntu distributions or\\
\verb|sudo yum install imagemagick| \\
for Fedora-compatible distributions).

Titles and axis legends must be part of the command(s) that generate
the graph or geometry window. For example, try this on a command line
\begin{center}
\verb|f(t):=sin(t);title="Graph of f on a period",|
\verb|labels=["t","f(t)"],legend=["cm","m"],|
\verb|plot(f(t),t=-pi..pi,legend="f")|
\end{center}

\subsubsection{2-d geometry}
Use the \verb|Geo->New figure 2d| menu to make a new 2-d geometry
session. This will open a new level with a ``subsession'' (group of
command lines) at the left,
a 2-d screen in the middle, and the controls at the right. If there
are parameters, they will be below the controls at the right.
The subsession is a list of command line levels. Evaluation of
a command line level will automatically reevaluate all levels
below (i.e. with a higher level number), therefore if you use
one level to define an object and modify the definition of this
object later, all objects depending on this object will be modified.
Modification of a level defining a point or a 1-d parameter can be
done interactively with the mouse either in the geometry screen or in the
parameter area (below the controls).

The geometry level has a mode, which affects how the mouse behaves.
\begin{itemize}
\item In all modes, right-clicking (or double-clicking) on an object
displays a dialog box so that you can easily modify the object
attributes (or delete the object).
\item In 2-d frame mode, draging the mouse will move the visualization window
and scrolling will zoom in/out.
\item In pointer mode, you can move an existing point, push the mouse 
near the point (when the mouse is near enough, it will show
a different cursor) and move the mouse, release it at the
new postion.
\item In point mode, the mouse can be used to define new
points or segments. To define a new
point, just click somewhere not near another point or segment. To
define a segment, press the mouse button at the beginning, move
the mouse, release the mouse at the end of the segment (if
you release the mouse out of the screen, the segment will be
cancelled). 
\item You can draw a few more complex object with the mouse using
the other modes, like triangles, circles, bisectors, .... 
\item You may also type Xcas commands (from the Geo menu)
in a command line at the left of the graphic. Each command may evaluate
to a graphic object and be displayed or to a non graphical object which
may be used later in the construction.
\end{itemize}
To erase a geometrical object, right-click on it, select Delete
or select its level in the
subsession at the left, and use the \verb|Edit->Delete selected levels|
menu or Backspace key.

For example, you can click three points with the mouse (in point mode),
this will create 3 levels with definitions for points A,B,C.
Then you can make the circle containing A, B, C either by selecting
the \verb|Circles->circumcircle| mode and click on A,B,C
or by entering in a command line the
\verb|circumcircle| command (menu \verb|Geo->Circles->circumcircle|)\\
\verb|c:=circumcircle(A,B,C)|

{\bf Remark: how to work with several figures in a tab}\\
When you create objects in a figure with a mouse, 
the name of the created object is
automatically chosen. This name is shared between figures in the same
tab so that they don't interfere if you have several figures in the
same tab. If you want to copy a part of a figure in another figure,
enter in a level of the target figure
the instruction {\tt eval([ ],1)} where you put inside the brackets
the names of the objects to be copied (e.g. {\tt eval([A,B,AB],1)}
will copy A, B and AB from a previous figure to the new figure). 
If you change the initial figure (where A,
etc. were defined), you must reevaluate the instruction in the target figure
to reflect the changes (press Enter in the first level of the target figure).

\subsubsection{Parameters}
Parameters are real values which can move in a fixed interval, e.g.
between -5 and +5. The real value may be changed by clicking on arrows
below the controls at the right of the geometrical
representation. There are two kinds of parameters: formal
and numeric parameters. Formal parameters have a formal name
which is used in all exact calculations, and a numeric value
which is used for all numeric evaluations (including graphical representation
of objects depending on the parameter). They can therefore be used
to make an analytical proof of a geometric theorem using 
the CAS engine while displaying
a figure. Numeric parameters always evaluate to their numeric value, 
they can
not be used to make an analytic proof (only to make conjectures 
like with most interactive geometry software).

To add a new parameter, use the {\tt Geo->Edit->Add parameter} menu.
For example, add a level with \verb|f(x):=a*x^2+b*x+c|
then define 3 parameters named a, b, c and add the command
\verb|plot(f(x))|, then move one of them to see the 
influence of each parameter
on the shape of the parabola.
You could also add a level \verb|mini:=solve(diff(f(x),x))| and
\verb|M:=point(mini[0],f(mini[0]))| to display the minimum.
Note  that the coordinates of M are displayed in terms of a and b
under the command line defining M at the left. You would have to add a level with
\verb|evalf(coordinates(M))| to see the current numeric value.

\subsubsection{3-d rendering.}
There are several options for 3-d graph rendering that can be
configured by clicking on the cfg button. First, you have to choose
between intrinsic colors or a scene viewed with 1 to 8 light spots, 
see below.

Then you can choose the $x$, $y$, $z$ range that will be visible.
The position of the eye visualizing the scene can be changed by
specifying a plane equation (normal to the vision direction). You
can also easily change the viewpoint outside of
the cfg configuration window either with the
mouse (push outside of the parallelepiped, drag the mouse
and release it at your option) or with the x,X, y,Y, z,Z keys to
rotate around the $x$, $y$ or $z$ axis. 

By using the r key or by selecting 
\verb|M->3-d->Rotate animation|,
you can get a visualization animation by rotation around the
origin. By default, it will rotate the viewpoint around the $z$ axis
but this is configurable from the \verb|cfg| button.  You can configure:
\begin{itemize}
\item \verb|Anim|: what will be animated, viewpoint if bit 8 is 1, 
spot number 7 to 0 if bit 7 to 0 are set to 1. For example, 255 will
move all spots leaving the viewpoint unchanged.
\item \verb|t|: interval between 2 redrawings.
\item \verb|n|: number of steps for a round trip.
\item \verb|x|, \verb|y|, \verb|z|: coordinates of the axes.
\item \verb|d|: normally 0, may be nonzero if you want also to move the
number of images of a programmed animation.
\end{itemize}
You can specify the axis of rotation in the command line defining the
scene by 
\begin{center}
\verb|gl_rotation_axis=[x,y,z]| 
\end{center}
where \verb|x,y,z|
are the coordinates of the axis of rotation.

Do not confuse visualization animation with programmed animation
obtained by instructions like \verb|animate, animate3d, animation|.

\subsubsection{3-d spots}
If you enable \verb|Lights| in the \verb|cfg| configuration of a 3-d
graph, the scene will be rendered by the OpenGL lighting
algorithm. You can enable 1 to 8 spots of light, each is configurable
(buttons \verb|L0| to \verb|L7|). 

Each spot has geometric properties: it may be a positional spot
if $w=1$ or a directional spot if $w=0$
\begin{itemize}
\item  A positional spot ($w=1$) is
at position $x,y,z$ and has a direction, defined by 
\verb|x->, y->, z->|. Light coming from a positional spot may
be attenuated, the attenuation may depend on the distance between
the spot and the object (with a quadratic, 
\verb|att0, att1, att2| in the spot configuration), and by the
cosine of the angle between the spot direction and the segment
spot-object to an exponent (\verb|exp| in the spot configuration).
In addition, the spot may light only inside a cone of angle
configurable by the \verb|cutoff| value (in degrees, use 180
for no cutoff or a value between 0 and 90). An additional attenuation
comes from the cosine of the angle incoming direction-normal to the
object, so that zenithal lighting is maximal.
\item A directional spot $w=0$ (emulating e.g. Sun light), does not
attenuate (except for the angle between the direction of the
spot and the normal). The direction is defined by $x,y,z$.
\end{itemize}

Objects from the scene may
have their own properties with respect to three kinds of light:
\begin{itemize}
\item ambient light,
emulating light coming from all directions and diffused
in all directions without attenuation, 
\item diffuse light,
emulating light coming from one direction and diffused in all directions
with an attenuation factor proportional to the cosine of the
angle of the incoming direction and the normal to the object,
\item specular light,
emulating light coming from one direction and reflected 
preferentially in the direction symmetric with respect to
the normal of the object.
\end{itemize}
In addition, an object may have intrinsic light, called emission.
Each property must be specified for 4 channels: r (red), b (blue)
and g (green) for the colors, and a (alpha) for transparency (if
you enable \verb|Blending|).
Object properties must be specified in their definition (on 
the command line):
\begin{itemize}
\item \verb|gl_material=[gl_front,gl_ambient,[r,g,b,a]]|\\ 
specifies the property for ambient light, use 
\verb|gl_diffuse|, \verb|gl_specular| or \verb|gl_emission|
instead of \verb|gl_ambient| for diffuse, specular or intrinsic
light. Default is 0.2 for ambient, 0.8 for diffuse and 0 for emission
on r,g,b channels and 1 on alpha channels.
\item
\verb|gl_material=[gl_front,gl_shininess,n]|\\
specifies the exponent $n$ that will be used for specular light. The
exponent is that of $\cos(\theta)$, where $\theta$ is the angle
between reflected incoming light and visualization direction (default
is $n=50$).
\item
\verb|gl_texture="filename"| specify an image filename that
will be mapped on a sphere or a polygon or a parametric
defined surface.
\end{itemize} 

Summing up, the formula for light on each channel is~:
\begin{eqnarray*}
 l &=&o_e+ o_a \sum_{j=1}^8 l_{a,j} + o_s \sum_{j=1}^8 l_{s,j}
 \cos(\gamma_j)^S + \\
 &  & +  o_d \sum_{j=1}^8 l_{d,j} 
\cos(\alpha_j)^{e_j} \cos(\beta_j) \frac{1}{a_{0,j}+a_{1,j}
  d_j+a_{2,j} d_j^2} 
\end{eqnarray*}
where~:
\begin{itemize}
\item $o_e,o_a,o_d,o_s$ are the material properties with respect to
 ambiant, diffuse, specular light, $S$ is the
``shininess'' (exponent for specular light),
\item $l_{a,j},l_{d,j},l_{s,j}$ are the properties of spot $j$ for
  ambiant, diffuse and specular light,
\item $a_{0,j}, a_{1,j}, a_{2,j}$ are attenuation coefficients
of spot $j$, replaced by 1,0,0 if $w=0$ (no attenuation),
\item $d_j$ is the distance of spot $j$ to the object,
\item $\alpha_j$ is the angle between the direction of spot $j$ and the
segment from spot $j$ and the object, except if $\alpha_j$ is
greater to the angle of cutoff of spot $j$ ($\alpha_j=\pi/2$), 
or if $w=0$ ($\alpha_j=0$, no attenuation for directionnal light),
\item $\beta_j$ is the angle between the segment spot $j$-object and
the normal to the object (if $w=0$, replace spot $j$-object by
the spot direction),
\item $\gamma_j$ is the angle between the reflection of the direction 
of the spot with respect to the normal of the object 
and the segment's object-eye.
\end{itemize}


\subsubsection{3-d geometry}
3-d geometry is based on the same principles as 2-d geometry: a
subsession at the left and the representation in the
middle. Mouse interaction outside the representation will
change the viewpoint, inside the representation it will depend
on the mode: draw a point, a segment, ... according to the mouse
plane displayed above (all mouse clicks are supposed to refer
to this plane, which is perpendicular to the visualisation axis).

\subsection{Spreadsheet}
Use the \verb|Spreadsheet->new spreadsheet| menu to add a spreadsheet level.
You will see a small configuration screen where you can change
the number of rows/columns, if an attached graph is displayed,
and also a variable name that you can use to get values
from command lines outside of the spreadsheet (using the matrix notation,
e.g. {\tt a[1,2]} will return the value of the cell at row 2, column 3
in {\tt xcas} syntax mode, or at row 1, column 2 in other syntax modes). 
You can modify the spreadsheet configuration by clicking on the
status line above the cells or with the 
spreadsheet \verb|File->variable name| menu later).
If you have chosen to display the graph inside the spreadsheet
configuration, it will dynamically
display all cells that evaluate to a graphic object. 
You may also display all the graphic
objects of the spreadsheet in a separate window, either in 2-d or in 3-d by 
clicking on the \framebox{2d} or \framebox{3d} button at the right of 
the spreadsheet menu.

A cell may contain
a fixed value (any valid Xcas object: integers, reals, symbols,
algebraic objects, strings, ...) or a formula depending on other
cells. The syntax for formulas is the same as for many spreadsheets, 
begin with an \verb|=|, then enter an algebraic expression where
other cells are represented using a symbol made of the column
name (e.g. A) and the row number (e.g. 1). When copying cells,
cell dependance in a formula is considered relative
unless you write a dollar (\$) sign before the column name or
row number. You can enter an area instead of a cell name, an area
is always a rectangular area, it is made of two cell names separated
by \verb|..| representing opposite vertices of the rectangular area.

To enter a new value or formula in a cell, click on the cell, then
enter your value or formula with the keyboard. As soon as you type
a normal character, the focus will move from the sheet to the
command line above the sheet. To help enter formula, you can use the mouse 
to select a cell or a cell area. Once the value or formula is correct
press the enter key to validate it. You can cancel your edit at
any time by clicking on a different cell.

To modify a cell value, click on the cell so that the value or formula
is displayed in the command line above, then click in the command line,
modify it and press Enter.

You can move in the sheet using the mouse or the direction keys, or
using the ``goto'' input value at the topleft of the sheet. Just enter
a cell name then enter to move there. If the cell does not exist,
the sheet will be enlarged to make it exist. You can also enter
an area in the ``goto'' input value to select this area without the
mouse. You must use the ``goto'' input value to make special
selections that are not connected areas, for example \verb|A1..B3,D,F| will
``select'' the subtable with rows 1 to 3 and columns A, B, D and F.
You can copy and paste a rectangular area with the mouse like this:
select the rectangular area with the mouse, click on the target cell,
click on the command line, type Ctrl-U to erase the previous
value then click on the middle button and type enter.

If you want to copy a cell formula to an area of the spreadsheet,
press at the right bottom part of the cell (when the mouse is on the
correct area of the cell, it will display a different cursor)
and move the mouse over the area you want to
copy, then release the mouse.
You can also copy cells using a menu item in the spreadsheet
\verb|Edit| menu, you can either copy to the left, or to the bottom,
or to a rectangular area previously selected with the mouse (in this
case the cell which was below the mouse when you pressed the button
will be copied to the rectangular area).

The spreadsheet can be saved, exported, imported, etc. using the
spreadsheet \verb|File| menu. The spreadsheet is also saved 
as part of the session (but not as a standalone spreadsheet)
when you save the whole session. When you load a session, the
spreadsheet will not be evaluated automatically, you must reevaluate
it with the \verb|reeval| button explicitly (or modify one cell).

Xcas does not provide filters for native binary spreadsheet formats,
but you can import spreadsheets from other software or data using
copy/paste or by saving them in the CSV (comma separated values) text format:
cells values must always be separated by the same character e.g. a comma, and
another character should be used to start a new row like a newline.
For example, you can select a spreadsheet area inside Open Office
and paste it inside Xcas.
If you want to import formulas instead of values,
use the Tools->Option menu of Open Office, find the spreadsheet Display item
and check Formulas. Select the area inside Open Office, then
paste at the same upper left cell position inside Xcas.
Alternatively you can save the spreadsheet in CSV format inside
Open Office with the \verb|File->Save as| 
menu item, then select CSV in the choosebox below. 
Then inside Xcas, use the \verb|File->Insert CSV| menu item,
select the file, change the separators if Xcas made a bad guess.
Note that Xcas must use row notation starting
at 1 instead of 0 for Open Office CSV formula import (this is the default).

Click on the status line or select inside the \verb|Edit| menu
for configuration and sheet modification operations. Configuration
operations should be self-explanatory, except for matrix fill cells : 
when entering a matrix in the command line, 
you can choose between two modes, either the matrix will fill
a rectangular area of cells, or it will fill only one cell.

The \verb|Stats|
menu is useful to make statistical plots in the corresponding graph.
When you make a graph using this menu, you will have to select
the data area where the statistic function will be applied (e.g.
a polygonplot on \verb|A1..B10|) and a target cell. The target
cell is a regular cell which was empty and will contain the formula
defining the graphic object that will be displayed. If you check the
value box, the graph will not change later, but if you don't check
the value box and modify
one data cell later, the graph will be updated like any cell 
depending on this data cell would be.

\subsection{Logo-like turtle}
To add a turtle session, use the \verb|Turtle-> new turtle| menu.
A logo level is made of three parts, the graphic display 
in the middle, a subsession of command lines at the left
and an editor at the right. Command lines are executed as in
a main session (command lines below an evaluated command
are not reevaluated unlike in a geometry session). The editor
records all commands.

In a logo level, you can pilot a turtle, giving it orders like forward
10 steps, turn left or right, etc. It will display a trace
using a pen with a defined color. 
You can use any language construction of Xcas to program the
turtle's moves. The current language uses French command names
(e.g. avance=move forward, etc.), the most important commandnames
are available by clicking on the buttonbar below the turtle screen
(check that the cursor is on a command line before clicking on
the button corresponding to the command you want to enter).
Most commands have a default argument (e.g. 10 for avance),
if you want to add parameters, do not add parentheses. Logo
commands use the same syntax as the \verb|return| statement,
if you want to specify priorities, put parenthesis around
the command and arguments, like this: \verb|(hasard 10)|.

\section{Session management}
Xcas may open different sessions, each with its own levels.
You can move levels inside a session or copy a level from one
session to another session.

To move one level in the session, press the mouse on its
level number and release it at the destination.
If you want to move some contiguous levels, click on
the first level, then shift-click on the last level (that
is, click on the mouse while pushing the shift key).
You can now move these levels by pressing the mouse in the
selected area and releasing it at its destination.
You can also copy these levels to another session by 
Ctrl-C/Ctrl-V or select/middle click.

You can delete the selected levels. You can paste the
selected levels elsewhere by clicking with the middle mouse button 
on another level number. You can
merge the selected levels in a single level.
You can also group the selected levels in a section and
give it a section name, sections may be folded or unfolded.

The \verb|File| menu of a session will let you save/export/load/import
a session in/from the current session. The \verb|Session| menu 
from the top menu will let you insert a new session.

\section{Configuration}
There are three main categories of configuration: CAS configuration
for the computer algebra commands, graphic configuration for the
default graphic options when creating a new graph, and general
configuration for other configuration items (like colors
or browser for help display, etc.). The first one is
accessed by the status button in the middle of the buttons line at the
top of the session.  All configuration items are also accessible using
the main  \verb|Cfg| menu.

\section{Customization}
You can alter part of the main menubar by editing a file named
{\tt xcasmenu}, either system-wide by the administrator
(in the Xcas {\tt doc/en} subdirectory), or only for you
(in the directory where you start \xcas). The same applies
to the Examples item, the filename being {\tt xcasex}.
The file {\tt keywords} may be used to translate Xcas command names
to another language.
\end{document}
